---
title: "Mixed Logit in R, with cross-language parity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed Logit in R, with cross-language parity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

This vignette walks through the R parity layer of the mixed-logit-choice-lab
project. The point is *cross-language reproducibility*: the same synthetic
data, the same model, comparable estimates from R's `mlogit` and the
project's Python implementation.

# 1. The shared data

Both languages read the same CSV, generated by the Python side with a
fixed seed:

```bash
# Run once from python/ to (re)generate the canonical dataset
python -m mixedlogit.export_csv
python scripts/export_results.py
```

This writes `r/data/synthetic_choices.csv` and the Python results JSON used
for parity comparison.

```{r}
source("../R/data.R")
df <- load_synthetic_data()
str(df)
```

The data has 12,000 rows: 500 individuals × 8 situations × 3 alternatives.
Long format, one row per (individual, situation, alternative).

# 2. Fitting MNL with `mlogit`

```{r, eval = requireNamespace("mlogit", quietly = TRUE)}
source("../R/estimate.R")
mnl <- fit_mnl_r(df)

# Coefficient table
data.frame(
  coef     = round(mnl$coefficients, 4),
  std_err  = round(mnl$std_errors, 4),
  t_value  = round(mnl$t_values, 3),
  row.names = mnl$coef_names
)
```

The MNL log-likelihood is `r if (exists("mnl")) round(mnl$loglik, 2) else "N/A"`,
McFadden's pseudo-R² is `r if (exists("mnl")) round(mnl$mcfadden_r2, 4) else "N/A"`.

# 3. Fitting MXL with `mlogit` random coefficients

```{r, eval = requireNamespace("mlogit", quietly = TRUE)}
mxl <- fit_mxl_r(df, n_draws = 200)

data.frame(
  coef     = round(mxl$coefficients, 4),
  std_err  = round(mxl$std_errors, 4),
  t_value  = round(mxl$t_values, 3),
  row.names = mxl$coef_names
)
```

# 4. Cross-language parity

The critical step: comparing R estimates against the Python results JSON.
These should match within simulation tolerance.

```{r, eval = file.exists("../data/python_mnl_results.json") && requireNamespace("mlogit", quietly = TRUE)}
source("../R/parity.R")
cmp_mnl <- compare_with_python(mnl, "../data/python_mnl_results.json")
cmp_mnl
```

```{r, eval = file.exists("../data/python_mxl_results.json") && requireNamespace("mlogit", quietly = TRUE)}
cmp_mxl <- compare_with_python(mxl, "../data/python_mxl_results.json")
cmp_mxl
```

The `gap_pct` column is the percentage difference between R and Python
estimates. Typical results: MNL gaps under 1% (the MNL likelihood is
deterministic, so the only differences come from the optimizer's
convergence tolerance), MXL gaps under 5% on means and under 10% on SDs
(MXL has simulation noise from the Halton draws).

# 5. Elasticities and the IIA failure

```{r, eval = requireNamespace("mlogit", quietly = TRUE)}
source("../R/elasticity.R")

# Differentiated design: Premium / Mid / Budget
design <- rbind(
  c(2.0,  1.0, 1.0),   # Premium
  c(1.0,  0.0, 1.0),   # Mid
  c(0.5, -1.0, 0.0)    # Budget
)
labels <- c("Premium", "Mid", "Budget")

mnl_E <- mnl_elasticities_r(mnl, design, alt_labels = labels)
mxl_E <- mxl_elasticities_r(mxl, design, alt_labels = labels, n_draws = 2000)

cat("MNL elasticity matrix:\n")
print(round(mnl_E$matrix, 3))
cat("\nMXL elasticity matrix:\n")
print(round(mxl_E$matrix, 3))
```

Look at the Premium column. Under MNL, Mid and Budget gain at exactly the
same rate when Premium raises price (the IIA signature). Under MXL, Mid
gains more — it's the closer substitute. This is the substitution realism
that justifies MXL for pricing decisions.

# 6. Willingness to pay

```{r, eval = requireNamespace("mlogit", quietly = TRUE)}
source("../R/wtp.R")
wtps <- compute_wtp_r(mxl, n_draws = 10000)
ranking <- feature_ranking_r(wtps, by = "median")
print(round(ranking, 3))
```

Quality has the higher median WTP (close to the true value of 0.667) and
a much wider IQR than brand recognition. **Quality is a segmentation
feature; brand is a broadly-shared baseline preference.** Same insight,
same numbers, in R as in Python.

# Summary

This vignette demonstrates the parity claim of the project: R and
Python produce the same estimates, the same elasticities, and the same
WTP rankings on the same synthetic data. The `gap_pct` columns above are
the proof.
